---
title: "RMarkdown_Sasidharan_et_al_2024_FEcol"
author: "Rohit Sasidharan"
date: "20-09-2024"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}

#### IMPORTING THE DATASETS
#Requires Rtools and devtools compatible with current versions of R and RStudio

#Installation
# install.packages("BiodiversityR")
# install.packages("ggplot2")
# install.packages("tidyverse")
# install.packages("reshape2")
# install.packages("lme4")
# install.packages("glmmTMB")
# install.packages("DHARMa")
# install.packages("emmeans")
# install.packages("vegan")

#### Packages needed
library(readxl) # read in excel file with function
# library(BiodiversityR) # can be used in place of vegan; also loads vegan
library(ggplot2) # for plots
library(readxl) # for reading excel files
library(readr) # read data in a tidy manner
library(here) # never write paths again
library(dplyr) # allows usage of %>% from magrittr and data wrangling
library(tidyr) # data wrangling
library(tidyverse) #text changes
library(reshape2) # for casting, melting, reshaping the data
library(lme4) #fitting LMMs
library(glmmTMB) #zero inflation models
library(DHARMa) #model checking
library(emmeans) #least sqare means
library(car) #Anova cmd 
library(stats) # various stat tests
library(vegan) # various stat tests


#pairwiseAdonis
library(devtools) # for installing pairwiseadonis
devtools::install("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
library(pairwiseAdonis) # for performing pairwise versions of adonis
#### datasheet 1: Terpenoids NMDS, Mantel and Varpart


data_1<- read_excel("SI_dataset.xlsx", 
    sheet = "Terpenoid_composition", col_types = c("text", 
        "text", "text", "text", "text", "text", 
        "skip", "skip", "numeric", 
        "skip"))


#melt the datasheet
md<-melt(data_1, id=c("Sample_Name","Plant","Genotype","Part","Chemotype","Compound"))
str(md) # view structure
## The relative area becomes the value

# transpose: cast the melted data
# cast(data, formula, function) 
#### IMPORTANT:  adds an extra number column : REMOVE THIS both in the next steps and the final version.

transdat<-dcast(md, Sample_Name+Plant+Chemotype+Genotype+Part~Compound,mean)

##Filtering the data files
data1f<-filter(transdat, Part %in% "Leaf")
data1l<-filter(transdat, Part %in% "Flower_head")

#### datasheet 2: Flower visitors
data_2 <- read_excel("SI_dataset.xlsx", 
    sheet = "Flower visitors_plantwise", 
    col_types = c("numeric", "numeric", "numeric", 
        "text", "text", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric"))

data_2_2021<-filter(data_2, Year %in% "2021")
data_2_2022<-filter(data_2, Year %in% "2022")

#### datasheet 3: Diversity counts
data_3 <- read_excel("SI_dataset.xlsx", 
    sheet = "Diversity_counts_plotwise", 
    col_types = c("numeric", "numeric", "numeric", 
        "text", "text", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric"))

#### datasheet 4: Germination
data_4 <- read_excel("SI_dataset.xlsx", 
    sheet = "Germination_expt", col_types = c("text", 
        "text", "text", "text", "numeric", 
        "text", "text", "text", "text", "text", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric"))
```

## R Markdown for Field Data

All data shown here is made avilable on Github: <https://github.com/R0hSa/FieldStudy_chemodiversity_2023>.
Rtools is required for the current version of R to run this script. 

## **1. Analyses of Terpenoid Profiles** : 

**NMDS of leaf and flower terpenoids**
The data provided here are transformed using Wisconsin square-root transformation. 
```{r terpenoids leaf}


#make community matrix - extract columns with abundance information

#Wisconsin square-root transformation
data_ws <- wisconsin(sqrt(transdat[,6:ncol(transdat)]))
data_ws_f<-wisconsin(sqrt(data1l[,6:ncol(data1l)]))
data_ws_l<-wisconsin(sqrt(data1f[,6:ncol(data1f)]))

#make community matrix
m_com = as.matrix(data_ws)
leafy_numbers<-as.matrix(data_ws_f)
flowery_numbers<-as.matrix(data_ws_l)


### NMDS

set.seed(500)
nmds = metaMDS(m_com, distance = "kulczynski")
nmds

plot(nmds)

#extract NMDS scores (x and y coordinates)

# use 
#data.scores = as.data.frame(scores(nmds)) #for R versions
#or change it to 
data.scores = as.data.frame(scores(nmds)$sites) #for R versions

#add columns to data frame 
data.scores$Sample_Name = transdat$Sample_Name
data.scores$Chemotype = transdat$Chemotype
data.scores$Part = transdat$Part

#PLOT

xx = ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) + 
  geom_point(size = 3, aes(colour = Chemotype,shape=Part))+ 
 #geom_text(aes(x=NMDS1,y=NMDS2,label=Sample_Name), size=3, alpha=1)+  ### to view sample labels
  theme(axis.text.y = element_text(colour = "black", size = 18, face = "bold"), 
        axis.text.x = element_text(colour = "black", face = "bold", size = 18), 
        legend.text = element_text(size = 19, face ="bold", colour ="black"), 
        legend.position = "right", axis.title.y = element_text(face = "bold", size = 16), 
        axis.title.x = element_text(face = "bold", size = 18, colour = "black"), 
        legend.title = element_text(size = 18, colour = "black", face = "bold"), 
        panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, linewidth  = 1.2),
        legend.key=element_blank()) + 
  labs(x = "NMDS1", colour = "Chemotype", y = "NMDS2",shape="Part")  + 
  scale_colour_manual(labels=c("Keto","BThu","AThu","Aacet","Myrox"),values = c("#FFC466","#77A8F5","#F9623E","#6ECFBD","#B481FF")) 

xx


```
**PERMANOVA**
The data provided here are transformed using Wiscon square-root transformation. 
```{r terpenoids PERMANOVA}
############ 
perm_result_f<-adonis2(flowery_numbers~Chemotype,permutations = 1000, method="kulczynski", data=data1f)
perm_result_f

perm_result_l<-adonis2(leafy_numbers~Chemotype,permutations = 1000, method="kulczynski", data=data1l)
perm_result_l

```
**pairwise PERMANOVA**
Ensure that the pairwiseadonis package is installed from https://github.com/pmartinezarbizu/pairwiseAdonis
```{r terpenoids pairwise PERMANOVA}
############ 
pairwise.adonis(
  dist.flower,
  flowery$Chemotype,
  sim.function = "vegdist",
  p.adjust.m = "BH",
  reduce = NULL,
  perm = 999
)

pairwise.adonis(
  dist.leaf,
  leafy$Chemotype,
  sim.function = "vegdist",
  p.adjust.m = "BH",
  reduce = NULL,
  perm = 999
)

```

**Mantel Test**
```{r terpenoids mantel}

#melt the data
md<-melt(data_1, id=c("Sample_Name","Plant","Genotype","Part","Chemotype","Compound","Area","Percent_Area"))


#transpose: cast the melted data
transdat3<-dcast(md, Sample_Name+Plant+Chemotype+Genotype+Part~Compound,mean)


#filter into leaf or flowerhead datasets
leafy<-filter(transdat3, Part %in% "Leaf")
leafy_numbers<-leafy[,6:ncol(leafy)]

flowery<-filter(transdat3, Part %in% "Flower_head")
flowery_numbers<-flowery[,6:ncol(flowery)]

#distance matrix: kulczynski
dist.leaf = vegdist(leafy_numbers, method = "kulczynski")
dist.flower = vegdist(flowery_numbers, method = "kulczynski")

#mantel
leaf_flower = mantel(dist.leaf, dist.flower, method = "spearman", permutations = 9999, na.rm = TRUE)
leaf_flower

```

**Variance partitioning**
```{r terpenoids varpart}

com2 = transdat3[,6:ncol(transdat3)]
m_com2 = as.matrix(com2)

# Two explanatory matrices -- Hellinger-transform Y
# Formula shortcut "~ ." means: use all variables in 'data'.
mod <- varpart(m_com2, ~Chemotype, ~Part, data=transdat3, transfo="hel")
mod
## e.g., http://127.0.0.1:10221/graphics/plot_zoom_png?width=1142&height=616

#If you want to plot it
# plot(mod, Xnames = c('Chemotype', 'Organ'), bg = c('deeppink3','seagreen3'), alpha = 80, # colour the circles
#        digits = 2, # only show 2 digits
#        cex = 1.5)
```

**ANOVA on variances**
```{r terpenoids varpart anova}

## rda on all fractions before a permutational anova can be performed
## a = Chemotype only, c = Part (organ) only, b = overlap between chemotype and part

# fraction [a]:
rda.Chemotype.Part <- rda (m_com2 ~ Chemotype + Condition (Part), data = transdat3)
# fraction [c]:
rda.Part.Chemotype <- rda (m_com2 ~ Part + Condition (Chemotype), data = transdat3)

# fractions [a+b+c]:
rda.all <- rda (m_com2 ~ Chemotype + Part, data = transdat3)
# fractions [a+b]:
rda.Chemotype <- rda (m_com2 ~ Chemotype, data = transdat3)
# fractions [b+c]:
rda.Part <- rda (m_com2 ~ Part, data = transdat3)


## The global model (fractions [a+b+c]):
anova (rda.all)

## Simple (marginal) effect of Chemotype (fraction [a+b]):
anova (rda.Chemotype)

## Simple (marginal) effect of Part (fraction [b+c]):
anova (rda.Part)

## Conditional (partial) effect of Chemotype (fraction [a]):
anova (rda.Chemotype.Part)

## Conditional (partial) effect of Part (fraction [c]):
anova (rda.Part.Chemotype)

#########

```

## **2. Flower Visitor Observations** : 
**Pollinator visits**
```{r pollinators}
modelp<-glmmTMB(Pollinator_counts~Plot_type*Chemotype+Day+Wind_speed+Year+Available_yellow_FHs+(1|Block)+(1|Block:BlockPlot)+(1|Genotype)+(1|Genotype:Plant_ID)+Temperature+Humidity,ziformula=~Day,family="nbinom1",data=data_2)

summary(modelp)

Anova(modelp,type="III",test="Chisq") 

```
**Florivore visits**
```{r florivores}
modelf<-glmmTMB(Florivore_counts~Chemotype*Plot_type+Day+Year+Available_FHs+(1|Block)+(1|Block:BlockPlot)+(1|Genotype)+(1|Genotype:Plant_ID)+Temperature+Humidity,ziformula=~Available_FHs,family="poisson",data=data_2)

summary(modelf)

Anova(modelf,type="III",test="Chisq") 

## for chemotype, pairwise comparisons
emmf <- lsmeans(modelf, ~ Chemotype) ## or emmeans
emmf
pairwise_comparisons_f <- pairs(emmf)
summary(pairwise_comparisons_f)
```
**Pollinator-florivore correlations (yearwise)**

```{r pollinator florivore corr}
#2021

modelpf_2021<-glmmTMB(Pollinator_counts~Florivore_counts+(1|Chemotype)+Plot_type+Day+Wind_speed+Available_yellow_FHs+(1|Block)+(1|Block:BlockPlot)+(1|Genotype)+(1|Genotype:Plant_ID)+Temperature+Humidity,ziformula=~Available_yellow_FHs,family="nbinom1",data=data_2_2021)

summary(modelpf_2021)

Anova(modelpf_2021,type="III",test="Chisq") 

#2022

modelpf_2022<-glmmTMB(Pollinator_counts~Florivore_counts+Available_yellow_FHs+Plot_type+(1|Chemotype)+Humidity+Temperature+Wind_speed+(1|Block)+(1|Block:BlockPlot)+(1|Genotype)+(1|Genotype:Plant_ID)+Day,ziformula=~Available_yellow_FHs,family="poisson",data=data_2_2022)

summary(modelpf_2022)

Anova(modelpf_2022,type="III",test="Chisq") 
```
**Plot level species diversity (richness)**
```{r richness}

#Pollinator species richness
modelpr<-lmer(Pollinator_richness~Plot_type+Day+Wind_speed+Year+Available_yellow_FHs+(1|Block)+(1|Block:BlockPlot)+Temperature+Humidity,data=data_3)

summary(modelpr)

Anova(modelpr,type="III",test="Chisq") 

#Florivore species richness
modelfr<-glmer(Florivore_richness~Plot_type+Day+Wind_speed+Year+Available_FHs+(1|Block)+(1|Block:BlockPlot)+Temperature+Humidity,family="poisson",data=data_3)


summary(modelfr)

Anova(modelfr,type="III",test="Chisq") 
```

## **3. Germination Data** : 
**Chemotype and Plot type effects**
```{r germination}
modelg<-lmer(Percentage_germination~Chemotype*Plot_type+Replicate+(1|Block:BlockPlot)+(1|Genotype)+(1|Genotype:Plant_ID),data=data_4) 

summary(modelg)

Anova(modelg,type="III",test="Chisq") 

#### for chemotype, pairwise comparisons
emm <- lsmeans(modelg, ~ Chemotype) ## or emmeans
emm
pairwise_comparisons <- pairs(emm)
summary(pairwise_comparisons)

```
**Correlations with pollinators or florivores**
```{r germination corr}
# Germination- Pollinator correlations
modelgp<-lmer(Percentage_germination~Pollinator_counts+Available_yellow_FHs+(1|Chemotype)+(1|Block:BlockPlot)+(1|Genotype)+(1|Genotype:Plant_ID),data=data_4)


summary(modelgp)

Anova(modelgp,type="III",test="Chisq") 

# Germination- florivore correlations
modelgf<-lmer(Percentage_germination~Florivore_counts+Plot_type+Available_FHs+(1|Chemotype)+(1|Block:BlockPlot)+(1|Genotype)+(1|Genotype:Plant_ID),data=data_4)

summary(modelgf)

Anova(modelgf,type="III",test="Chisq") 

```
## **---------------------------------------THE END----------------------------------------------** 